Negative Selection in the 20th Century

### 2020-06-19 why-negative-selection.Rmd
More work on article

# Open questions/TODO

* If we weight with the GHS, and plot the means of figure 1 using this w
  weighting, most things stay the same - but the change in EA is reversed! EA
  increases over time. Could this be because the sample includes many
  relatively educated older people? If so, then mean effects are *overestimated*
  by figure 1. 
* Reweight other regressions (e.g. the omnibus regression in the appendix)
* Could be useful: f.2139 - age first had sex (includes "never had sex" 
  which may explain some of the many NAs for f.2141, num sex partners)
* Look at geography, esp. in siblings regressions.
* control for age in n_partners regressions.

@@

We run bivariate regressions on two dependent variables:

* *siblings*, the number of full siblings in the respondent's sib (including
  himself or herself).
* The number of *children* ever born to/fathered by the respondent.

```{r calc-regs-all}

res_all_wide <- res_all %>% 
  filter(reg.type == "raw") %>% 
  select(dep.var, term, estimate) %>% 
  pivot_wider(names_from = dep.var, values_from = estimate)

n_pgs  <- as.double(nrow(res_all_wide))

prop_gens <- res_all_wide[["N siblings"]]/res_all_wide[["N children"]]
median_prop_gens <- median(prop_gens[prop_gens > 0])
```

Figure \@ref(fig:plot-basic-regs) shows effect sizes of a one-standard
deviation shift in each polygenic score.

Estimates are broadly consistent across generations. However, effect sizes are 
much larger for *siblings* than for *children* 
regressions. Among consistently-signed estimates, at the median, the proportion between
effect sizes for siblings and for children is `r 100 * median_prop_gens` per cent.

```{r plot-basic-regs, fig.cap = "Effects of polygenic scores on number of siblings/children."}

n_regs <- as.double(nrow(res_all))
res_all %>% 
      filter(reg.type == "raw") %>% 
      mutate(
        term    = pretty_names(term),
        term    = fct_reorder(term, estimate, order_abs(1)),
        dep.var = fct_inorder(dep.var) 
      ) %>% 
      ggplot(aes(y = term, x = estimate, fill = dep.var, 
        colour = p.value < 0.05/{{n_regs}})) +
      my_vline + 
      pval_geom_point() +
      my_labs() +
      pval_scale() +
      my_fill_scale(n = 2, guide = guide_legend(title = "")) +
      my_theme()

# Subgroups

We next examine how different subgroups of the population contribute to natural 
selection. These analyses have two goals: to get a sense of how ascertainment bias
could affect our basic results, and to learn more about the causes of natural 
selection in the UK population. Because we have more data on respondents than on
their parents, we focus on *children* regressions.

## Sex, income and education

Figure \@ref(fig:plot-educ-level) splits respondents up by education levels.
Selection effects are typically larger and more significant for those who left
school before 16.  Figure \@ref(fig:plot-income) splits respondents by household
income category. A very similar pattern holds, with selection effects being
larger for those in the poorest income category.

@@

The UK Biobank sample overrepresents highly educated people and high-income
households. This suggests that statistics from the sample may underestimate
the level of natural selection in the population.

## Age at first live birth

@@

Age at first live birth correlates with number of children ever born. Figure
\@ref(fig:plot-age-flb-cross) shows *children* regressions estimated separately
for each tercile of age at first live birth. Several effects are strikingly
different across terciles. ADHD and MDD are selected for amongst the youngest
third of mothers, but selected against among the oldest two-thirds. Educational
attainment is selected for among the oldest two-thirds of mothers, but is not
significantly selected among the youngest third. Similarly, several PGS of body
measurements are selected against only among older mothers. The correlation
between effect sizes for the youngest and oldest terciles is `r cor_age_flb`.
This is the first sign that selection may work in different directions across
different subgroups.

Figure \@ref(fig:plot-n-partners) splits males and females by lifetime number of
sexual partners, at the median value of 3. Again, selection effects are reversed
across the groups. Among men and women with more than 3 sexual partners, for
example, EA is selected against; among those with 3 or fewer sexual partners, it
is selected for. Selection is broadly negative (positive) among those 
with more (less) partners.

Lastly Figure \@ref(fig:plot-with-partner) splits respondents up by whether they
were living with a spouse or partner at the time of interview. Effects are 
generally the same sign in both groups, but are much larger among those not
living with a spouse or partner.

@@

# Accounting for ascertainment bias

Our results are subject to two kinds of ascertainment bias. First, Biobank is
not a random sample of the UK population: for example, it overrepresents
educated people and high income households. Second, sampling parents of
respondents inherently overweights parents who have many children. For instance,
parents of three children will have three times more children as respondents in
UK Biobank than parents of one child, on average. Parents of no children will,
by definition, not be represented at all.

drake::loadd(res_wt_ghs_weights)
drake::loadd(res_wt_flb_weights)
drake::loadd(res_wt_msoa_weights)
drake::loadd(res_wt_age_qual_weights)
drake::loadd(res_sibs_parent_weights)
drake::loadd(res_all)

res_wt_cb_wide <- res_wt_combined %>% 
                    tidyr::pivot_wider(
                      id_cols     = term,
                      names_from  = Weights,
                      values_from = estimate
                    ) %>% 
                    rowwise() %>% 
                    mutate(
                      consistent = all(c_across(-term) > 0) ||
                        all(c_across(-term) < 0)
                    ) %>% 
                    ungroup() %>% 
                    left_join(
                      res_wt_combined %>% 
                        filter(Weights == "None") %>% 
                        select(term, p.value)
                    )

rel_sizes <- res_wt_cb_wide %>% 
                filter(
                  consistent, 
                  p.value < 0.05/length(score_names)
                ) %>% 
                mutate(
                  across(Geographical:`Age/Qual/FLB`, ~ .x/None)
                ) %>% 
                select(- consistent, - None, - p.value) %>% 
                mutate(
                  term = pretty_names(term)
                )

rel_sizes_summary <- rel_sizes %>% 
                summarize(
                  across(Geographical:`Age/Qual/FLB`, ~ c(mean(.x), median(.x)))
                ) %>% 
                mutate(
                  term = c("Mean", "Median")
                )

* Geographical weighting by census MSOA, age and presence/absence of a partner;
* Weighting by age and highest educational qualification;
* Weighting by age, highest qualification, and age at first live birth, using
  GHS data.

Figure \@ref(fig:plot-res-wt) shows the results. Effect sizes are significantly
larger in absolute value. Table \@ref(tab:tbl-res-wt) gives effect sizes as a 
proportion of unweighted effect sizes.

```{r tbl-res-wt}
rel_sizes %>% 
      arrange(desc(`Age/Qual/FLB`)) %>% 
      bind_rows(rel_sizes_summary) %>% 
      rename(PGS = term) %>% 
      as_hux() %>% 
      set_caption(
        "Weighted effect sizes as a proportion of unweighted"
      ) %>% 
      insert_row("", "Weighting", fill = "") %>% 
      merge_cells(1, 2:4) %>% 
      set_align(1, 2, "centre") %>% 
      set_header_rows(1, TRUE) %>% 
      theme_compact() %>% 
      style_headers(bold = TRUE) %>% 
      set_bottom_border(1, 2:4) %>% 
      set_bottom_border(2, everywhere) %>% 
      set_background_color(final(2), everywhere, "grey80") %>% 
      add_footnote(
        "Only consistent and significant (when unweighted) estimates are shown",
        font_size   = 8, 
        top_padding = 2
      )

res_wt_combined %>% 
  mutate(
    Weights  = fct_relevel(Weights, "Geographical", "Age/Qualification",
                "Age/Qual/FLB", "None"),
    term     = pretty_names(term),
    term     = fct_reorder(term, estimate, order_abs(4))
  ) %>% 
  ggplot(aes(x = estimate, y = term, shape = Weights, colour = Weights)) +
  geom_point(size = 2) +
  my_vline +
  scale_colour_manual(values = c(
    "None"              = "grey25", 
    "Geographical"      = "darkgreen",
    "Age/Qualification" = "chocolate4",
    "Age/Qual/FLB"      = "midnightblue"
  )) +
  scale_shape_manual(values = c("None" = "square", "Geographical" = "circle",
        "Age/Qualification" = "circle", "Age/Qual/FLB" = "circle"))

We then calculate the 
proportions of respondents within each sibling group size, based on the respondent
weights, and use this to create weights for the sibling regressions.

```{r calc-flb-weights}

drake::loadd(res_wt_flb_weights)
drake::loadd(res_sex)
res_flb_all <- left_join(
        res_wt_flb_weights, 
        res_sex %>% filter(sex=="Female") %>% select(-subset, -sex), 
        by = "term", 
        suffix = c(".wt", ".nowt")
      ) %>% 
      mutate(prop_estimate = estimate.wt/estimate.nowt)

prop_wt_nowt <- mean(res_flb_all$prop_estimate)
pct_change_score <- function (score) {
  100 * (res_flb_all$prop_estimate[res_flb_all$term == score] - 1)
}
```

We also weighted data for women with children only, by age, education and age 
at first live birth.  Figure \@ref(fig:plot-flb-weights) 
shows the results. There 
are large changes for some variables, including EA3 
(`r pct_change_score("EA3_excl_23andMe_UK")` per cent), cognitive
ability (`r pct_change_score("cognitive_ability")` per cent) and
waist-hip ratio (WHR; `r pct_change_score("whr_combined")` percent).

Overall, these results suggest that previous work may have underestimated the level of
selection occurring in the population. [TODO: who?]

```{r plot-flb-weights, fig.cap = "Effect sizes on number of children, female respondents weighted by age at first live birth"}

bind_rows(
        Weighted   = res_wt_flb_weights, 
        Unweighted = res_sex %>% filter(sex == "Female") %>% select(-subset, -sex),
        .id        = "Regression"
      ) %>% 
      standard_ggplot(Regression, n_regs = 66)
```

## Mechanisms behind natural selection

Why are selection effects *oppositely* signed in certain subgroups -- for
example, in older and younger mothers (by age at first live birth) or in those
with more and fewer lifetime sexual partners? A possible answer is given by the
economic theory of fertility [TODO cite Becker 1960]. According to this,
increases in potential earned income affect fertility via two opposing channels.
There is an "income effect" by which children become more affordable, like any
other good. Since childrearing has a cost in time, there is also a substitution
effect: the opportunity cost of childrearing increases if one's market wage is
higher. The income effect would lead higher earners to have more children. The
substitution effect would lead higher earners to have fewer children. It is
often assumed that the substitution effect will dominate for lone parents, since
they have less opportunity to share childcare responsibilities, while the income
effect will dominate for couples who are able to reap the gains from
specialization.

If so, then genetic characteristics which affect one's earnings potential in the
labour market may lead to opposing effects on fertility. Genetic
variants which improve one's earnings may increase fertility among couples, but
decrease fertility among single parents.

This explanation predicts that the strength of polygenic scores' effects on
fertility will correlate with the strength of their relationship with earnings.
We measure the correlation of each PGS with household income, as well as with
education level, a phenotypic variable which is likely to predict earnings. We
then relate these measurements to effect sizes on number of children.
Across scores, there is a clear negative relationship between effect on
fertility, and the correlation with income and/or education (Figure \@ref(fig:plot-cors-income-educ).

```{r plot-cors-income-educ, fig.cap="Effect sizes on number of children and correlations with income and education"}

drake::loadd(res_all)
cors_income <- cor(famhist[c(score_names)], famhist$income_cat, use = "pairwise")
# TODO: use age_left_edu when this includes college
cors_educ <-   cor(famhist[c(score_names)], famhist$edu_qual, use = "pairwise")

dfr <- data.frame(
      type = rep(c("Cor. with income", "Cor. with educ"), each = 33),
      cor = c(cors_income[,1], cors_educ[,1]),
      effect_size = rep(effect_size, 2)
)

ggplot(dfr, aes(effect_size, cor, colour = type)) + 
      geom_point() +
      facet_wrap(vars(type)) + 
      my_vline + 
      my_hline +
      labs(x = "Effect on n children", y = "Correlation") +
      scale_colour_manual(values = c("darkblue", "darkgreen")) +
      theme(legend.position = "none")

@@

Figure \@ref(fig:plot-age-flb) shows the results of *children* regressions for 
women only,
controlling for age at first live birth. Effect sizes are greatly reduced. Partly,
this is because the regressions exclude childless women. However,
in `r opp_signed` out of `r length(score_names)` cases, effect sizes actually
change sign with the controls. The correlation between effect sizes
controlling for age at first live birth, and raw effect sizes, is `r cor_raw_flb`.

@@

>>>>

```
# Open questions/TODO

* If we weight with the GHS, and plot the means of figure 1 using this w
  weighting, most things stay the same - but the change in EA is reversed! EA
  increases over time. Could this be because the sample includes many
  relatively educated older people? If so, then mean effects are *overestimated*
  by figure 1. 
* Reweight other regressions (e.g. the omnibus regression in the appendix)
* Look at geography, esp. in siblings regressions.
* control for age in n_partners regressions.
* can we weight men by age at first child, using GHS? If not, may need
  to plot the age_flb weighted regressions separately, against unweighted
  women-only regressions.
* Can we find representative data for weights by n_partners?
```

@@

These estimates may suffer from ascertainment bias within the UK Biobank sample.
Respondents are higher income and better educated than the UK population, and
they may also differ on other unobserved characteristics.

To examine the effects of sampling, we regress respondents' polygenic
scores on their number of children:

\[
y_i = \alpha + \beta\mathrm{PGS}_i + \varepsilon_i
\]

The effect size $\beta$ measures the strength of natural selection within the
sample (or a given subgroup). In fact, since polygenic scores are normalized to
mean 0, $\beta$ is the expected polygenic score among children of the sample.

Figures \@ref(fig:plot-educ-level) and \@ref(fig:plot-income) plot effect sizes
for each polygenic score among different subgroups of the sample. Figure
\@ref(fig:plot-educ-level) groups respondents by age of completing full-time
education. Figure \@ref(fig:plot-income) groups respondents by household income.
Effect sizes are larger and more significant for the lowest income category, and
for the lowest education category. Note that the overall effect is not a simple
average of the effect among the different subgroups, because polygenic scores 
may also shift respondents between the subgroups. For example, a high 
PGS for educational attainment will increase the age at which a respondent 
leaves school.

@@

When we split respondents up by other demographic variables, we get a more
surprising result: the direction of natural selection is different among
different subgroups.

@@

Figure \@ref(fig:plot-age-flb-cross) shows effect sizes estimated separately for
each tercile of age at first live birth. Several effects are strikingly
different across terciles. ADHD and MDD are selected for amongst the youngest
third of mothers, but selected against among the oldest two-thirds. Educational
attainment is selected for among the oldest two-thirds of mothers, but is not
significantly selected among the youngest third. Similarly, several PGS of body
measurements are selected against only among older mothers. The correlation
between effect sizes for the youngest and oldest terciles is `r cor_age_flb`.

Figure \@ref(fig:plot-n-partners) splits males and females by lifetime number of
sexual partners, at the median value of 3. Again, selection effects are reversed
across the groups. Among men and women with more than 3 sexual partners, for
example, educational attainment is selected against; among those with 3 or fewer
sexual partners, it is selected for.

Figure \@ref(fig:plot-with-partner) splits respondents up by whether they
were living with a spouse or partner at the time of interview. Effects are 
generally the same sign in both groups, but are much larger among those not
living with a spouse or partner.

@@

@@

Figure \@ref(fig:plot-age-flb) shows the results of regressions for women only,
controlling for age at first live birth. Effect sizes are greatly reduced.
This could be because the regressions exclude childless women. However, in 
`r opp_signed` out of `r length(score_names)` cases, effects actually change
sign when controls are added. The correlation between effect sizes controlling for age
at first live birth, and raw effect sizes, is `r cor_raw_flb`.

@@

## Weighted regressions

drake::loadd(res_wt_ghs_weights)
drake::loadd(res_wt_flb_weights)
drake::loadd(res_wt_msoa_weights)
drake::loadd(res_wt_age_qual_weights)
drake::loadd(res_sibs_parent_weights)
drake::loadd(res_all)
drake::loadd(res_sex)

res_wt_cb_wide <- res_wt_combined %>% 
                    tidyr::pivot_wider(
                      id_cols     = term,
                      names_from  = Weights,
                      values_from = estimate
                    ) %>% 
                    rowwise() %>% 
                    mutate(
                      consistent = all(c_across(-term) > 0) ||
                        all(c_across(-term) < 0)
                    ) %>% 
                    ungroup() %>% 
                    left_join(
                      res_wt_combined %>% 
                        filter(Weights == "None") %>% 
                        select(term, p.value),
                      by = "term"
                    ) %>% 
                    left_join(
                      res_sex %>% 
                        filter(sex == "Female") %>% 
                        select(term, estimate_females = estimate),
                      by = "term"
                    )

rel_sizes <- res_wt_cb_wide %>% 
                filter(
                  consistent, 
                  p.value < 0.05/length(score_names)
                ) %>% 
                mutate(
                  Geographical = Geographical/None,
                  `Age/Qualification` = `Age/Qualification`/None,
                  `Age/Qual/FLB` = `Age/Qual/FLB`/estimate_females
                ) %>% 
                select(-consistent, -None, -p.value, -estimate_females) %>% 
                mutate(
                  term = pretty_names(term)
                )

rel_sizes_summary <- rel_sizes %>% 
                summarize(
                  across(-term, ~ c(mean(.x), median(.x)))
                ) %>% 
                mutate(
                  term = c("Mean", "Median")
                )

* Geographical weighting by census MSOA, age and presence/absence of a partner;
* Weighting by age and highest educational qualification;
* Weighting by age, highest qualification, and age at first live birth, using
  GHS data (women only).

Figure \@ref(fig:plot-res-wt) plots effect sizes among the entire sample, using
these weighting schemes. We include unweighted regressions for comparison.
Weighted effect sizes are significantly larger in absolute value. Table
\@ref(tab:tbl-res-wt) gives effect sizes as a proportion of unweighted effect
sizes for the census-based weightings.

Effect sizes are reasonably consistent across the geographical and
age/qualification weights. But they are increased again when we weight by age at
first live birth.This suggests that estimates might be increased still more by
weighting on other relevant demographic variables. Overall, these results
suggest that previous work may have underestimated the level of selection
occurring in the population. [TODO: who?]

```{r tbl-res-wt}

rel_sizes %>% 
      arrange(desc(Geographical)) %>% 
      bind_rows(rel_sizes_summary) %>% 
      rename(PGS = term) %>% 
      as_hux() %>% 
      insert_row("", "Weighting", fill = "") %>% 
      set_caption(
        "Weighted effect sizes as a proportion of unweighted"
      ) %>% 
      merge_cells(1, 2:4) %>% 
      set_align(1, 2, "centre") %>% 
      set_header_rows(1, TRUE) %>% 
      theme_compact() %>% 
      style_headers(bold = TRUE) %>% 
      set_bottom_border(1, 2:4) %>% 
      set_bottom_border(2, everywhere) %>% 
      set_italic(final(2), 1) %>% 
      set_font_size(8) %>% 
      set_width(0.7) %>% # ensures line wrapping below
      set_col_width(c(.34, .22, .22, .22)) %>% 
      add_footnote(paste0(
        "Only consistent and significant (when unweighted) estimates ",
        "are shown.\n",
        "Age/Qual/FLB as a proportion of unweighted regressions ",
        "for females only."),
        font_size    = 6, 
        top_padding  = 0,
        left_padding = 0
      )

n_regs <- length(score_names)
res_wt_combined %>% 
  mutate(
    Weights  = fct_relevel(Weights, "Geographical", "Age/Qualification",
                "Age/Qual/FLB", "None"),
    term     = pretty_names(term),
    term     = fct_reorder(term, estimate, order_abs(1))
  ) %>% 
  ggplot(aes(x = estimate, y = term, shape = Weights, fill = Weights, colour = p.value < 0.05/{{n_regs}})) +
  geom_point(size = 1.5, stroke = 0.5) +
  my_vline +
  pval_scale() +
  scale_fill_manual(values = c(
    "None"              = "grey15", 
    "Geographical"      = "dodgerblue4",
    "Age/Qualification" = "steelblue2",
    "Age/Qual/FLB"      = "darkseagreen4"
  )) +
  scale_shape_manual(values = c(
    "None" = "square filled", 
    "Geographical" = "circle filled",
    "Age/Qualification" = "circle filled", 
    "Age/Qual/FLB" = "circle filled"
  ))

## Change over time

The UK Biobank data contains information on respondents' number of siblings.
Since respondents' polygenic scores are equal in expectation to their parents' 
mean scores, we can use this to look at natural selection in the parents'
generation. The dependent variable is now *number of siblings* (including
the respondent him- or herself).

The parents' generation has a second source of ascertainment bias, beyond the
sampling bias of the survey. Sampling parents of respondents inherently
overweights parents who have many children. For instance, parents of three
children will have, on average, three times more children as respondents in UK
Biobank than parents of one child. Parents of no children will
not be represented at all, by definition.

To compensate for this, we reweight our preferred weightings by the inverse of
*number of siblings*.

TODO: 
- weight regressions using "best effort" weights
- look at change over time to confirm your "no change" hypothesis
- look at siblings, again using "best" weights

## Mechanisms behind modern natural selection

Why do selection effects have *opposite* signs across certain subgroups -- for
example, across older and younger mothers by age at first live birth, or across those
with more and fewer lifetime sexual partners? A possible answer is given by the
economic theory of fertility [TODO cite Becker 1960]. According to this,
increases in potential earned income affect fertility via two opposing channels.
There is an "income effect" by which children become more affordable, like any
other good. Since childrearing has a cost in time, there is also a substitution
effect: the opportunity cost of childrearing increases if one's market wage is
higher. The income effect would lead higher earners to have more children. The
substitution effect would lead higher earners to have fewer children. It is
often assumed that the substitution effect will dominate for lone parents, since
they have less opportunity to share childcare responsibilities, while the income
effect will dominate for couples who are able to reap the gains from
specialization.

If so, then genetic characteristics which affect one's earnings potential in the
labour market may lead to opposing effects on fertility. Genetic
variants which improve one's earnings may increase fertility among couples, but
decrease fertility among single parents. They may also decrease the likelihood
of becoming a young mother, if this is correlated with single motherhood. Lastly,
if lifetime number of sexual partners predicts single parent versus 
two-parent status, the theory could also explain the differently-signed
effects across people with more and fewer sexual partners.

This explanation predicts that the strength of polygenic scores' effects on
fertility will correlate with the strength of their relationship with earnings.
We measure the correlation of each PGS with household income, as well as with
education level, a phenotype which predicts earnings. We
then relate these measurements to effect sizes on number of children.
There is a clear negative relationship between a score's effect on
fertility, and its correlation with income or education (Figure \@ref(fig:plot-cors-income-educ)). If we control for age at first live birth,
the relationship changes sign and becomes positive.

```{r plot-cors-income-educ, fig.cap="Effect sizes on number of children, with and without controls for age at first live birth, by correlations with income and education"}

drake::loadd(res_all)
drake::loadd(res_age_flb)

cors_income <- cor(famhist[score_names], famhist$income_cat, use = "pairwise")
# TODO: use age_left_edu when this includes college
cors_educ <-   cor(famhist[score_names], famhist$edu_qual, use = "pairwise")

effect_size_flb <-  res_age_flb %>% 
                      filter(term != "age_flb") %>% 
                      pull(estimate)

dfr <- data.frame(
        Correlation  = rep(c("Income", "Education"), each = 33, 2),
        cor          = rep(c(cors_income[,1], cors_educ[,1]), 2),
        effect_size  = c(rep(effect_size, 2), rep(effect_size_flb, 2)),
        Controls     = rep(c("None", 
                        "Age at first live birth"), each = 66)
      )

dfr %>% 
      mutate(
        Controls = fct_relevel(Controls, "None")
      ) %>% 
      ggplot(aes(effect_size, cor, colour = Correlation)) + 
        geom_point() +
        facet_grid(Correlation ~ Controls, labeller = label_both) + 
        my_vline + 
        my_hline +
        labs(x = "Effect on number of children", y = "Correlation") +
        scale_colour_brewer(type = "qual", palette = 2) +
        theme(legend.position = "none")
